### 2022-06-20 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@63a51b8ed8b53a85690ceaf73d894e2b7187b884 🚀

Spillover risk is multidimensional and complicated. Within a pool of
competent hosts, it is driven by a multiplicity of factors (Plowright et
al. 2017). Although proxies for the local richness of hosts are commonly
analysed (Anthony et al. 2017, Ruiz-Aravena et al. 2022), there is an
argument to be made that species who are not (known as) competent hosts
of a specific virus genus may not factor into this (Plowright et al.
2015), calling for species-level information. This is especially true as
competence data increases predictive power when the taxonomic scope of
hosts of a viral family increases (Becker et al. 2020, Mull et al.
2022). Similarly, host species who share viruses at different rates
should be weighted accordingly (Albery et al. 2020). In mammals, key
functional traits (for which phylogeny is a reasonable proxy) are
determinants of the spillover potential of their viruses (Olival et al.
2017); these include, notably, body mass, life history, diet, and use
for human-occupied spaces (Albery et al. 2022). Finally, especially when
the pool of potential hosts spans the entire globe, there may be local
host pools that are highly unique; not having been observed in other
locations, these can act on the overall risk either by providing novel
contact opportunities, reflecting unique host-environment combinations
(Engering et al. 2013), or facilitating rapid evolutionary changes in
specialism of their pathogens (Agosta et al. 2010). In the specific case
of generalist pathogens (which is the case for many viruses in the
\emph{Betacoronavirus} genus, see \emph{e.g.} MacLean et al. 2021),
there is conceptual and empirical support to the idea that these
community-level mechanisms are even more important in driving the
overall risk (Power and Mitchell 2004). In this paper, we examine the
biogeographic structure of bats-betacoronaviruses associations, based on
a curated dataset of all confirmed bat hosts of betacoronaviruses, by
adopting an analysis framework grounded in concepts from the Geographic
Mosaic Theory of Coevolution (GMTC; Thompson 2005). The GMTC posits that
coevolutionary processes between pairs (Thompson 1994) or complexes
(Janzen 1980) of species are structured in space by the rippling effects
of abiotic conditions onto evolutionary mechansism, resulting in
spatially fragmented evolutionary dynamics, coupled only by
dispersal-related processes (Gomulkiewicz et al. 2000). In turn, these
spatially fragmented processes can lead taxonomically homogeneous
systems (in our case, the bats-betacoronaviruses complex) to have
different structure and dynamics of large spatial extents (see
\emph{e.g.} Price 2002).

The GMTC predicts a spatial fragmentation of coevolutionary dynamics
under the joint action of three processes (see notably Gomulkiewicz et
al. 2007), which all have the potential to act on outbreak potential,
pathogen tramission, and disease virulence (Parratt et al. 2016, Turner
et al. 2021). First hot/cold spots of coevolution can appear when the
intensity of \emph{interaction} (in terms of reciprocal fitness
consequences) varies spatially, because of \emph{e.g.} partial range
overlap between organisms (Nuismer et al. 2003); hot/cold spots are
notoriously difficult to identify in nature (Laine 2005), especially
when the structure of range overlap is complex, and the reciprocal
fitness effets are relatively low. Both of these conditions are met in
the bats-coronaviruses complex. Second, GMTC supposes the existence of
selection mosaics, wherein the intensity of \emph{selection} (in terms
of selective consequences of fitness effects) varies across space; the
strength of reciprocal selection responds to the biotic complexity of
the community (locally diverse hosts and viruses are more biotically
complex; Thrall et al. 2007) \emph{and} to the local favorability of the
environment (Hochberg and Baalen 1998). In a system with near global
distribution and variations in relative richness, like the
bats-betacoronaviruses complex, the GMTC predicts the establishment of
many different coevolutionary trajectories. The final GMCT process is
trait remixing, under which changes in coevolutionary dynamics happen
due to the arrival (or departure) of functional traits, through changes
in community composition due to invasions, meta-community dynamics, and
disperal; adopting an host-centric perspective, high viral sharing would
expose the host immune system to multiple new viral traits, thereby
changing the selection landscape for all viruses.

We turn the processes on the GMTC into definitions of spillover risk
from viruses to hosts (focusing on the bats-betacoronavirus complex),
with a specific emphasis on the potential to create independant
coevolutionary dynamics (and therefore spatial fragmentation in the
risk) through heterogeneity. Our components of risk are (i) the
phylogenetic diversity of hosts, representing the diversification
potential of the system (hotspots, selection mosaics), under the
assumption that bat diversification is not primarily driven by viruses
(Gorbunova et al. 2020); (ii) the local uniqueness of the bat community,
representing the potential for viruses to be exposed to novel host
traits in unusual combinations (cold spots, trait remixing); finally,
(iii) viral sharing among hosts, representing the potential for hosts to
be exposed to complex and functionaly diverse communities (trait
remixing, biotic complexity). We turn these predicates into a spatially
explicit additive mapping of zoonotic risk components, revealing the
extreme heterogeneity of risk at the global scale. Explicitely framing
the notion of spillover risk based on propositions from the GMTC (which
is to say, based on a framework linking interactions between species to
change within species) is a novel idea, that should be relatively
general. Indeed, it only assumes the action of well described
evolutionary mechanisms. The benefit of this approach is to provide the
potential for a more dynamic and nuanced understanding of risk: not only
on ecological timescales, but also by providing clues about which areas
can change over micro-evolutionary timescales. This provides a way to
look at spatial structure by accounting for more notions than species
richness/similarity, but also a way to identify spatial areas of higher
risk.

@@

\leavevmode\hypertarget{ref-Becker2020InfInt}{}%
Becker, D. J. et al. 2020. Beyond Infection: Integrating Competence into
Reservoir Host Prediction. - Trends in Ecology \& Evolution 35:
1062--1065.

@@

\leavevmode\hypertarget{ref-Engering2013PatHos}{}%
Engering, A. et al. 2013. Pathogen--host--environment interplay and
disease emergence. - Emerging Microbes \& Infections 2: e5.

@@

\leavevmode\hypertarget{ref-MacLean2021NatSel}{}%
MacLean, O. A. et al. 2021. Natural selection in the evolution of
SARS-CoV-2 in bats created a generalist virus and highly capable human
pathogen. - PLOS Biology 19: e3001115.

@@

\leavevmode\hypertarget{ref-Mull2022VirIso}{}%
Mull, N. et al. 2022. Virus isolation data improve host predictions for
New World rodent orthohantaviruses. - Journal of Animal Ecology in
press.

@@

\leavevmode\hypertarget{ref-Plowright2015EcoDyn}{}%
Plowright, R. K. et al. 2015. Ecological dynamics of emerging bat virus
spillover. - Proceedings of the Royal Society B: Biological Sciences
282: 20142124.

>>>>

Spillover risk is complex and, within a pool of susceptible wildlife
hosts, is driven by a multiplicity of factors (Plowright et al. 2017).
Although proxies for the local richness of hosts are commonly analysed
(Anthony et al. 2017, Ruiz-Aravena et al. 2022), these approaches
deliberately oversimplify interspecific heterogeneity in immunity,
behavior, and contact with humans. At the coarsest level, some species
play a more central role in the host-virus network because of key
functional traits (for which phylogeny is a reasonable proxy) (Albery et
al. 2020); these include, notably, body mass, life history, diet, and
use for human-occupied spaces Olival et al. (2017). Global maps of
spillover risk often struggle to distill these features into
interpretable risk maps, and overlook highly-unique host pools that
allow for the rapid evolution of highly divergent viruses (Agosta et al.
2010). In the case of generalist pathogens like betacoronaviruses, there
is conceptual and empirical support to the idea that these
community-level mechanisms are even more important in driving the
overall risk (Power and Mitchell 2004). In this paper, we examine the
biogeographic structure of bat-betacoronavirus associations by adopting
an analysis framework grounded in concepts from the Geographic Mosaic
Theory of Coevolution (GMTC; Thompson 2005). The GMTC posits that
coevolutionary processes between pairs (Thompson 1994) or complexes
(Janzen 1980) of species are structured in space by the rippling effects
of abiotic conditions onto evolutionary mechanism, resulting in
spatially fragmented evolutionary dynamics, coupled only by
dispersal-related processes (Gomulkiewicz et al. 2000). In turn, these
spatially fragmented processes can lead taxonomically homogeneous
systems to have different structure and dynamics over large spatial
extents (see \emph{e.g.} Price 2002).

The GMTC predicts a spatial fragmentation of coevolutionary dynamics
under the joint action of three processes (see notably Gomulkiewicz et
al. 2007), which all have the potential to act on outbreak potential,
pathogen tramission, and disease virulence (Parratt et al. 2016, Turner
et al. 2021). First hot- and coldspots of coevolution can appear when
the intensity of \emph{interaction} (in terms of reciprocal fitness
consequences) varies spatially, because of \emph{e.g.} partial range
overlap between organisms (Nuismer et al. 2003). Hot- and coldspots are
notoriously difficult to identify in nature (Laine 2005), especially
when the structure of range overlap is complex, and the reciprocal
fitness effects are relatively low---both of which conditions are met in
the bat-betacoronavirus complex. Second, the GMTC supposes the existence
of selection mosaics, wherein the intensity of \emph{selection} varies
across space; the strength of reciprocal selection responds to the
biotic complexity of the community (locally diverse hosts and viruses
are more biotically complex; Thrall et al. 2007) \emph{and} to the local
favorability of the environment (Hochberg and Baalen 1998). In a system
with near global distribution and variations in relative richness, like
the bat-betacoronavirus complex, the GMTC predicts the establishment of
many different coevolutionary trajectories (and resulting viral
regimes). The final GMCT process is trait remixing, under which changes
in coevolutionary dynamics happen due to the arrival (or departure) of
functional traits, through changes in community composition due to
invasions, meta-community dynamics, and disperal; these changes are
likely to happen in overlap zones between ``long branches'' of
coevolution.

We turn the processes on the GMTC into definitions of spillover risk
from viruses to hosts (focusing on the bats-betacoronavirus complex),
with a specific emphasis on the potential to create independant
coevolutionary dynamics (and therefore spatial fragmentation in the
risk) through heterogeneity. Our components of risk are (i) the
phylogenetic diversity of hosts, representing the diversification
potential of the system (hotspots, selection mosaics), under the
assumption that bat diversification is not primarily driven by viruses
(Gorbunova et al. 2020); (ii) viral sharing among hosts, representing
levels of viral exchange within host communities that may act as a
buffer against viral specialization in any given host (i.e., high
sharing creates coldspots of selective pressures from any given host
species); and (iii) the local uniqueness of the bat community,
representing the potential for viruses to be exposed to novel host
traits in unusual combinations (trait remixing). We turn these features
into a spatially-explicit additive mapping of zoonotic risk components,
revealing the extreme heterogeneity of risk at the global scale.
Explicitely framing the notion of spillover risk based on propositions
from the GMTC (which is to say, based on a framework linking
interactions between species to change within species) is a novel idea,
that should be relatively general. Indeed, it only assumes the action of
well described evolutionary mechanisms. The benefit of this approach is
to provide the potential for a more dynamic and nuanced understanding of
risk: not only on ecological timescales, but also by providing clues
about which areas can change over micro-evolutionary timescales. This
provides a way to look at spatial structure by accounting for more
notions than species richness/similarity, but also a way to identify
spatial areas of higher risk.

@@

@@

@@

@@

@@
